---
title: "Ordination"
author: "Aidan Fauth"
format: html
editor: visual
---
## Introduction

This document will be my preliminary data analysis for one of my questions regarding the salamander data:
What does the salamander community structure look like in both the transitional and mature forest?

To address this question, I will use ordination analysis.

## Set up

```{r}
rm(list=ls())
library(tidyverse)
library(here)
library(ggfortify)
library(vegan) # needed for ordination
```


## Load in the data

```{r}
sabund <- read.csv(here("Data", "Fauth_updated_salamander_abundance.csv"))
scount <- read.csv(here("Data", "Fauth_updated_salamander_count.csv"))
```

#### Assign needed data types:

Species

```{r}
scount$Species <- as.factor(scount$Species)
levels(scount$Species)
```

Site_Type

```{r}
scount$Site_Type <- as.factor(scount$Site_Type)
levels(scount$Site_Type)
```


## Plot

To get a sense of the data let's look at a bar graph of the frequencies of each species in each forest.

First we need count data to make the plot. Use the sum of the Count variable to get an actual total for each group:

```{r}
fs_count <- scount |> group_by(Site_Type,Species) |> summarise(total = sum(Count))
head(fs_count,12)
```

Clearly the red backs are way more common.

```{r}
ggplot(fs_count, aes(x = Site_Type, y = total, fill = Species))+
     geom_bar(stat = "identity", position = "dodge")
```

Again, the red backed salamander is the most common, and there were more found in the transitional forest. That is true of the other species as well, but there numbers are so small its hard to see a clear difference, other than the spotted salamander (a clear preference for the transitional forest as well.)




