---
title: "Data Exploration"
author: "Aidan Fauth"
format: html
editor: visual
---

## Set Up

```{r}
rm(list=ls())
library(tidyverse)
library(here)
library(ggfortify)
```

## Load in the datasets

```{r}
scount <- read.csv(here("Data", "Fauth_cleaned_salamander_data.csv")) # the first csv with the counts data
sabund <- read.csv(here("Data", "Fauth_salamander_richness_abundance.csv")) # file with the abundace data
```

Now, let's take a look at each of the data frames:

Start with scount:

```{r}
glimpse(scount)
```

In this data frame we have 1,734 rows of data with 10 variables. 
There are 5 numeric variables, 4 should be factors (the Count variable is a binary variable coded as 1s and 0s) and the Date should be changed to a date datatype.


sabund:

We expect to see a similar structure just with two variables changed (per the metadata file)

```{r}
glimpse(sabund)
```

This data frame has 1,560 entries (rows) and 10 variables (with Species and Count switched for Abundance and Richness.) 
There are 7 numeric variables, 2 variables should be factors, and there is the same date variable that needs to be converted to date data type.

Note: The total number of entries in each dataset are not the same, and therefore, I cannot merge the two files into one (which would greatly simplify my eventual analysis.)





